Increasing Productivity Amid Stable Nutrient Regimes in Rhode Island Lakes and Reservoirs

### 2019-07-29 manuscript/manuscript.Rmd
added one sentence to trends section

As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while total nutrients and the TN:TP ratio are all relatively stable.(Is is worth discussing that the current nutrient load is stable but probably high due to previous loadings?)  These stable nutrient regimes may be partly explained by efforts to curb nutrient loadings (e.g., [EQIP at USDA (QK will provide), ]).  Additionally, these results point to the fact that productivity is driven by processes operating at different scales.  For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [Boyer et al. 2002].  Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as catchment percent impervious and lake morphology can also impact temperature [Nelson & Palmer, 2007]. Thus, differences in regional and state level trends are driven by complex processes.

In addition to the annualized trends we address with this study, there are other trends that may be of interest.  For example, what are the trends for water quality at finer temporal scales such as monthly or seasonal trends?  Anecdotal evidence in Rhode Island [SHOULD REFERENCE THIS SOMEHOW - let's ask Elizabeth] points to greater increases in temperature earlier and later in the growing season and suggests a lengthening of the growing season. Furthermore, preliminary analysis of the URIWW data back this up with mean temperature for May 1993 to May 1995 cooler by nearly a degree than mean temperature for May 2014 through May 2016.  Additionally, are trends influenced by the current state of a given waterbody?  For instance, are oligotrophic lakes showing stronger trends than eutrophic lakes or are all lakes showing similar trends regardless of current trophic status? (Is this worth discussing further? There is at least one paper that could be mentioned.) These questions are beyond the scope of this study, but all warrant further, careful investigation.

>>>>

As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while total nutrients and the TN:TP ratio are all relatively stable. These stable nutrient regimes may be partly explained by efforts to curb nutrient loadings (e.g., [EQIP at USDA (QK will provide), ]). Although stable, there are likely sufficient nutrients present, as a result of eutrophication, to allow for increases in productivity. Additionally, these results point to the fact that productivity is driven by processes operating at different scales.  For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [Boyer et al. 2002].  Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as catchment percent impervious and lake morphology can also impact temperature [Nelson & Palmer, 2007]. Thus, differences in regional and state level trends are driven by complex processes.

In addition to the annualized trends we address with this study, there are other trends that may be of interest.  For example, what are the trends for water quality at finer temporal scales such as monthly or seasonal trends?  Anecdotal evidence in Rhode Island [SHOULD REFERENCE THIS SOMEHOW - let's ask Elizabeth] points to greater increases in temperature earlier and later in the growing season and suggests a lengthening of the growing season. Furthermore, preliminary analysis of the URIWW data back this up with mean temperature for May 1993 to May 1995 cooler by nearly a degree than mean temperature for May 2014 through May 2016.  Additionally, are trends influenced by the current state of a given waterbody?  For instance, are oligotrophic lakes showing stronger trends than eutrophic lakes or are all lakes showing similar trends regardless of current trophic status?  These questions are beyond the scope of this study, but all warrant further, careful investigation.
